Mixing Time Estimation in Reversible Markov Chains from a Single Sample Path

### 2015-08-09 nips2015/rebuttal.txt

We thank the referees for the helpful comments. We'll incorporate the suggestions in a revision of the paper, especially where more detailed explanations are needed.

The following two concerns were common to multiple reviews.

Tightness of bounds: We don't claim tightness of the bounds (we did not attempt to optimize the constants) but our results give the first non-trivial (upper and lower) bounds in this challenging problem. In particular, the estimator is based on a less than straighforward, indirect approach while avoiding slow rates (as explained at the beginning of Section 4) that builds on advanced perturbation results from the theory of stochastic matrices. Our work is motivated by practical applications where fully data-dependent intervals were sought but not actually achieved. Our results are a first step towards a practical solution: we demonstrate feasibility and also prove limitations on what is possible, both add significantly to our knowledge.

@@

5. Combining Theorems 1 and 2 with a standard argument yields the lower bound in the introduction. We will include the details in the revision.

6. Regarding Theorem 1: Note that "c" is an absolute constant in the theorem, not something that can be chosen arbitrarily. In particular, one value that works in the theorem is c=1/4.

There were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here. We need to change the "1/6" in the theorem statement and proof to "1/8". Furthermore, in the proof, \bar\pi should be taken in (0,1/4), and we should set \varepsilon := \bar\pi. Observe that \pi^{(2)} = (1/(1+2\varepsilon),2\varepsilon/(1+2\varepsilon)). Thus, \pi_\star \geq \bar\pi in both P^{(1)} and P^{(2)}. To guarantee |\hat\gamma_\star - \gamma_\star| < 1/8, it must be possible to distinguish the two Markov chains. The rest of the proof is correct as is.

7. The conclusion of Theorem 2 follows using the second-moment method (e.g., Paley-Zygmund) and the moment bounds that, with probability > 1/4, the earliest time in which a sample path visits all d states is at least c d log(d) / \bar\gamma.  This implies that an estimator using a shorter sample path cannot distinguish P^0 from P^i for some i (with probability > 1/4), and hence must have large estimation error. We'll include these details in the revision.

Minor comment 3: We respectfully disagree. The definition of a confidence interval does not distinguish intervals based on what is known, which itself may depend on the application (in specific applications additional knowledge if often available).

### 2015-09-12 arxiv-v2/abstract.tex
issues about relaxation time

This article provides the first procedure for computing a fully data-dependent interval that traps the mixing time $\tmix$ of a finite reversible ergodic Markov chain at a prescribed confidence level.  The interval is computed from a single finite-length sample path from the Markov chain, and does not require the knowledge of any parameters of the chain.  This stands in contrast to previous approaches, which either only provide point estimates, or require a reset mechanism, or additional prior knowledge.  The width of the interval converges to zero at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.  Upper and lower bounds are given on the number of samples required to achieve constant-factor multiplicative accuracy.  The lower bounds indicate that, unless further restrictions are placed on the chain, no procedure can achieve this accuracy level before seeing each state at least $\Omega(\tmix)$ times on the average.  Finally, future directions of research are identified.

>>>>

This article provides the first procedure for computing a fully
data-dependent interval that traps the mixing time
$t_{\text{mix}}$ of a finite
reversible ergodic Markov chain at a prescribed confidence level.  The
interval is computed from a single finite-length sample path from the
Markov chain, and does not require the knowledge of any parameters of
the chain.  This stands in contrast to previous approaches, which
either only provide point estimates, or require a reset mechanism, or
additional prior knowledge.
The interval is constructed around the relaxation time
$t_{\text{relax}}$, which is strongly related to the mixing time, and
the width of the interval converges to zero
at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.
Upper and lower bounds are given on the number of samples required to
achieve constant-factor multiplicative accuracy.  The lower bounds
indicate that, unless further restrictions are placed on the chain, no
procedure can achieve this accuracy level before seeing each state at
least $\Omega(t_{\text{relax}})$ times on the average.  Finally, future
directions of research are identified.
